---
title: "google_earth_engine"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Uses
- fire index, run-off, wetness, greeness/deforestation, migration
- pulls data from a lot of satellites
- a lot of the imagery can be pre-processed
- Depending on your question, what satellite to use?
  - spatial resolution (0.5m - 1km)
  - spectral resolution
  - temporal resolution (often trades off with spatial resolution)
    - landsat is 30 m, get an image every 2 weeks
  - $ vs. open source
   - everything in GEE is open source, but LiDAR is pricey
  
#Application
*GEE Explorer*
- you can do quick checks of what data are available, you can bring in your own shapefiles, too
- this is quick and dirty stuff

*GEE Code Editor*
- the better option
- middle: Run applies to everything in that window, comment it out if you don't want it to run
- left: assets is a way to manage maps between scripts, scripts includes example scripts
- right: inspector is a way to investigate pixels
  - get link allows you to send a URL to collaborators
- GEE data catalogue to decide what raster data sets to use

*Java script*
- "//" to comment
- "=" for the get
- purple font is built-in GGE functions
- () used to add parameters
- [] for a string or list
- ; to close an argument
- "var" tells it you are giving a variable name

*Practice*
- a composite is built of several satellite images taken of one spot, so when pulled in you lable it "imagecollection"
- any vector pulled into GEE is a geometry
- any raster is called an image, need a start and end date for the composite, tell it the bounds are the vector geometery that we set
- calculate median value of every pixel in the region of interest
- select the 3 bands you want to show and tell us where you want the map to zoom
- when you've made the thing, you can save it in GEE (asset) or in googledrive to then use in R

pull in the tif from GEE
```{r}
library(raster)

#upload GEE tiff
landsat<- raster("") #get filename from googledrive
plot(landsat)

```

  
   